The Unicode Cookbook \linebreak for Linguists

### 2015-11-15 book/chapters/ipa_background.tex
revision of first two sections of IPA introduction

\section{Brief history}

@@

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems.

Although the IPA began its life as a pedagogical tool, from its earliest days the 
Association aimed to provide ``a separate sign for each distinctive sound; 
that is, for each sound which, being used instead of another, in the same 
language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
sounds became later known as \textsc{phonemes} and the IPA has developed historically 
into a notational devise with a strictly segmented phonemic view. A phoneme is an 
abstract theoretical notion derived from an acoustic signal as produced by speakers 
in the real world. Therefore the IPA contains a number of theoretical assumptions 
about speech and how to analyze speech in written form.

Phonetic analysis is based on two premises: (i) that it is possible to describe
the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered
discrete segments, and, (ii) that each segment can be characterized by an
articulatory target.\footnote{A purely phonetic description is only derivable
from instrumental data in high quality sound recordings. Once spoken language
data are segmented, phonological consideration inextricably play a roll in
transcription. In other words, phonetic observations beyond quantitative
acoustic analysis are always made in terms of some phonological framework.}
Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67-68]{International1989report}:

@@

\noindent and in the \textit{Handbook of the International Phonetic Association} \citep[18]{IPA1993}:

@@

\noindent Although the IPA provides symbols to unambiguously represent phonemes,
it also aims to represent phonetic details. Since phonetic detail could
potentially include anything, e.g. something like `deep voice', the IPA
restricts phonetic detail to linguistically relevant aspects of speech. These
principles and conventions for using the IPA are outlined in the Association's
Handbook.

\section{Principles}

IPA transcription has essentially two parts. The first is a text containing IPA 
symbols and the second is a set of conventions (rules) for interpreting those 
symbols (and their combinations). The IPA is designed to meet practical linguistic 
needs and is used to transcribe the phonetic or phonological structure of languages. 
It is also used increasingly as a foreign language learning tool, as a standard 
pronunciation guide and as a tool for creating practical orthographies of 
previously unwritten languages.

The current construction and use of the IPA are guided by principles outlined 
in the \textit{Handbook of the International Phonetic Association: A guide 
to the use of the International Phonetic Alphabet} \citep[159]{IPA1999}, henceforth 
HB. The use of the symbols to represent a language's phonological system is guided 
by the principle of contrast; where two words are distinguishable by phonetic contrast, 
those contrasts should be transcribed with different symbols (graphemes not diacritics). 
Allophonic distinction falls under the rubric of diacritically-distinguished symbols, 
e.g. [stop̚] vs [spʰot]. In sum:

\begin{itemize}
	\item different symbols (without diacritics) should be used whenever a language employs two contrastive sounds
	\item when two sounds in a language are not known to be contrastive, the same symbol should be used to represent these sounds; however, diacritics may be used to distinguish such sounds when necessary
	\item diacritics cannot be dispensed with entirely, so the Association recommends to limit their use to:
	\begin{itemize}
		\item denoting length, stress and pitch
		\item representing minute shades of sounds
		\item obviating the design of a (large) number of new symbols when a single diacritic suffices (e.g. nasalized vowels, aspirated stops)
	\end{itemize}
\end{itemize}

\noindent Thus, an IPA transcription \textit{always} consists of ``a set of 
symbols and a set of conventions for their interpretation''.

In systematic transcription (as opposed to impressionistic), there is a division 
between phonemic and allophonic transcription. The terms phoneme and allophone 
contain theoretical baggage, but the basic goal of a phonemic transcription is to 
distinguish all words in a language with the minimal number of transcription 
symbols \citep[19]{Abercrombie1964}. Allophonic transcription uses a broader 
set of distinct symbols to describe systematic allophonic differences in sounds 
in words. These two systematic transcriptions are related to each other by a 
set of conventions in the IPA tradition, so that they can be converted between 
one and another.

An IPA transcription is connected to a speech event by a set of conventions. 
A phonetic (or impressionistic) transcription may use the conventions implicit 
in the IPA chart, i.e. the transcriber can indicate that the phonetic value of 
<ŋ͡m> is a simultaneous labial and velar closure which is voiced and contains 
nasal airflow.

A phonemic transcription includes the conventions of a particular language's 
phonological rules. These rules determine the realization of that language's 
phonemes. However, there can be different systems of phonemic transcription 
for the same variety of a language. The differences may result from the fact 
that more than one phonetic symbol may be appropriate for a phoneme (see Section 
\ref{}). Or the differences may be due to different phonemic analyses, e.g.\ 
Standard German's vowel system is arguably contrastive in length or tenseness.

An important principle of the IPA is that different representations resulting 
from different symbols and different analyses are in line with the IPA's aims. 
In other words, the IPA does not provide phonological analyses for specific 
languages and the IPA does not define a single ``correct'' transcription system. 
Rather, the IPA aims to provide a resource that allows users to express any 
analysis so that it is widely understood. Thus the IPA suits many linguists' 
needs because:

\begin{itemize}
	\item it is intended to be a set of symbols for representing all possible sounds in the world's (spoken) languages
	\item its chart has a linguistic basis (specifically a phonological bias) rather than a general phonetic notation scheme
	\item its symbols can be used to represent distinctive feature combinations\footnote{Although the chart uses traditional manner and place of articulation labels, the symbols can nevertheless represent any defined bundle of features, binary or otherwise, to define phonetic dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge (a common denominator of phonological ``facts'')
\end{itemize}

Several styles of transcription with IPA are possible and the HB illustrates 
these and notes that they are all valid (see the 29 languages and their transcriptions 
in the original and initial \textit{Illustrations of the IPA} \citep[41--154]{IPA2007}). 
Therefore, there are different but equivalent transcriptions, or as \cite[64]{Ladefoged1990a} 
captures it, ``Perhaps now that the Association has been explicit in its eclectic 
approach, outsiders to the Association will no longer speak of \textit{the} IPA 
transcription of a given phenomenon, as if there were only one approved style.''.

Clearly not all phoneticians agree (or will likely ever agree of course) on all 
aspects of the IPA or on transcription practices. As noted above, there have 
been several revisions in the IPA's long history, but the current version (2005) 
is strikingly similar to the 1926 version. In 1989 an IPA revision convention 
was held in Kiel, Germany. As per other revisions, there was expansion and revisions 
to phonetic symbols in the IPA chart. Notably the marking of tone was extended 
with the addition of a second system for marking linguistic tones (Chao tones). 
Importantly, however, this was the first revision to address issues of 
computational representations for the IPA -- the principles above of which 
have had several ramifications that make the interoperability of electronic 
linguistic data extremely difficult.

\section{Computational representation}

>>>>

% ==========================
\section{Brief history}
\label{IPAhistory}
% ==========================

@@

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textit{Handbook of the International Phonetic
Association: A guide to the use of the International Phonetic Alphabet}
\citep[159]{IPA1999}, henceforth simply called \textit{Handbook}.

Today, the IPA is designed to meet practical linguistic needs and is used to
transcribe the phonetic or phonological structure of languages. It is also used
increasingly as a foreign language learning tool, as a standard pronunciation
guide and as a tool for creating practical orthographies of previously unwritten
languages. The IPA suits many linguists' needs because:

\begin{itemize}

	\item it is intended to be a set of symbols for representing all possible
       sounds in the world's (spoken) languages
	\item its chart has a linguistic basis (specifically a phonological bias)
       rather than a general phonetic notation scheme
	\item its symbols can be used to represent distinctive feature
       combinations\footnote{Although the chart uses traditional manner and
       place of articulation labels, the symbols can nevertheless represent any
       defined bundle of features, binary or otherwise, to define phonetic
       dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge
       (a common denominator of phonological ``facts'')

\end{itemize}

Several styles of transcription with IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid (see the 29 languages and their
transcriptions in the original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}). Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``Perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all phoneticians agree (or will likely ever agree of course) on all 
aspects of the IPA or on transcription practices. As noted above, there have 
been several revisions in the IPA's long history, but the current version (2005) 
is strikingly similar to the 1926 version, showing the viability of the IPA as a 
common standard for linguistic transcription.

% ==========================
\section{Premises}
\label{IPApremises}
% ==========================

Any IPA transcription is based on two premises: (i) that it is possible to
describe the acoustic speech signal (i.e.\ sound waves) in terms of sequentially
ordered discrete segments, and, (ii) that each segment can be characterized by
an articulatory target.\footnote{A purely phonetic description is only derivable
from instrumental data in high quality sound recordings. Once spoken language
data are segmented, phonological consideration inextricably play a roll in
transcription. In other words, phonetic observations beyond quantitative
acoustic analysis are always made in terms of some phonological framework.}
Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67-68]{International1989report}:

@@

\noindent and in the \textit{Handbook} \citep[18]{IPA1999}:

@@

Although the IPA provides symbols to unambiguously represent phonemes, it also
aims to represent phonetic details. Since phonetic detail could potentially
include anything, e.g. something like `deep voice', the IPA restricts phonetic
detail to linguistically relevant aspects of speech. From its earliest days the
Association aimed to provide ``a separate sign for each distinctive sound; that
is, for each sound which, being used instead of another, in the same language,
can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive sounds
became later known as \textsc{phonemes} and the IPA has developed historically
into a notational devise with a strictly segmental phonemic view. A phoneme is
an abstract theoretical notion derived from an acoustic signal as produced by
speakers in the real world. Therefore the IPA contains a number of theoretical
assumptions about speech and how to analyze speech in written form.

% ==========================
\section{Principles}
\label{IPAprinciples}
% ==========================

Transcription has essentially two parts. The first is a text containing IPA 
symbols and the second is a set of conventions for interpreting those 
symbols (and their combinations).

First, the symbols of IPA crucially distinguish between letter-like symbols and
diacritics.\footnote{see Section~\ref{pitfall-different-notions-of-diacritics} for a more
detailed discussion about the notion of diacritics} The use of the letter-like
symbols to represent a language's sounds is guided by the principle of contrast;
where two words are distinguishable by phonemic contrast, those contrasts should
be transcribed with different letter symbols (not diacritics). Allophonic
distinction falls under the rubric of diacritically-distinguished symbols, e.g.
the difference in English between and unrealized /p/ in [stop̚] vs. an aspirated
/p/ in [spʰot] should be marked with a diacritic. In sum:

\begin{itemize}

	\item Different letter-like symbols should be used whenever
          a language employs two sound contrastively.
	\item When two sounds in a language are not known to be contrastive, the same
          symbol should be used to represent these sounds. Diacritics may
          be used to distinguish different articulations when necessary.
\end{itemize}

\noindent Yet, in some situations diacritics can be used to mark contrasts. The
\textit{Handbook} recommends to limit the use of phonemic diacritics to the
following situations:

\begin{itemize}

 	\item denoting length, stress and pitch;
	\item obviating the design of a (large) number of new symbols when a 
		  single diacritic suffices (e.g. nasalized vowels, aspirated stops).

\end{itemize}

The interpretation of the IPA symbols in specific usage is not trivial. Although
the articulatory properties of the IPA symbols themselves are rather succinctly
summarised by the normative description in the \textit{Handbook}, it is common
in practical applications that the transcribed symbols do not precisely
represent the phonetic value of the sounds that they represent. So an IPA symbol
<t> in one transcription is not always the same as an IPA <t> in another
transcription (or even within a single transcription). The interpretation of any
particular <t> is mostly a language-specific convention (or even
resource-specific and possibly even context-specific), which -- unfortunately --
is in most cases not made explicit by users of the IPA.

There are different reasons for this difficulty to interpret IPA symbols,
both officially part of the IPA specifications. Basically, the IPA allows for
both a `narrow' phonetic transcription and a `broad' phonological transcription.
An important principle of the IPA is that different representations resulting
from different underlying analyses are allowed. Because the IPA does not provide
phonological analyses for specific languages, the IPA does not define a single
``correct'' transcription system. Rather, the IPA aims to provide a resource
that allows users to express any analysis so that it is widely understood.

A `narrow' phonetic transcription may freely use the all symbols in the IPA
chart with direct reference to their phonetic value, i.e. the transcriber can
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound is
a simultaneous labial and velar closure which is voiced and contains nasal
airflow, independently of the phonemic status of this sound in the language in
question.

% examples from Wells \url{https://www.phon.ucl.ac.uk/home/wells/transcription-ELL.pdf}

In contrast, the basic goal of a phonemic transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

Note also that there can be different systems of phonemic transcription for the
same variety of a language, so two different resources on the `same' language
might use different symbols that represent the same sound. The differences may
result from the fact that more than one phonetic symbol may be appropriate for a
phoneme, or the differences may be due to different phonemic analyses, e.g.\
Standard German's vowel system is arguably contrastive in length, tenseness or
vowel quality. Finally, even within a single phonemic transcription a specific
symbol can have different realizations because of allophonic variation which 
might not be explicitly transcribed.

In sum, there are three different reasons why phonemically-based IPA 
transcription is difficult to interpret:

\begin{itemize}

   \item A symbol used represents the phonemic status, and not necessarily the
         precise phonetic realization. So, different transcriptions might use 
         the same symbol for different underlying sounds.
   \item The symbol that is used for a specific phoneme is not necessarily
         unique, so different transcriptions might use different symbols for the
         same underlying sound.
   \item Allophonic variation can be disregarded in phonemic transcription, so
         the same symbol within a single transcription can represent different
         sounds.

\end{itemize}

Ideally, all such implicit conventions of a phonemic transcription would be
explicitly codified. This could very well be performed by using an orthography
profile (see Chapter~\ref{orthography-profiles}), linking the selected phonemic 
transcription symbols to `narrow' phonetic transcriptions, possibly also including 
specifications of contextual interpretation.

% ==========================
\section{Computational representation}
% ==========================

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (Chao tones). Importantly, however, this was
the first revision to address issues of computational representations for the
IPA -- the principles above of which have had several ramifications that make
the interoperability of electronic linguistic data extremely difficult.

### 2015-11-15 book/chapters/ipa_meets_unicode.tex
revision of first two sections of IPA introduction

   \item According to the IPA, <ʰ> can be used both as <\dia{02B0}> for
         post-aspiration (following the base character) and as
         <\diareverse{02B0}> for pre-aspiration (preceding the base character).
         Note that there is yet another homoglyph, namely the
         \textsc{combining latin small letter h} at \uni{036A}. We propose not
         to use this combining diacritical mark, but to standardize on Unicode
         modifier letters for all superscript letters in IPA.

>>>>

   \item According to the IPA, <ʰ> can be used both as <\dia{02B0}> for
         post-aspiration (following the base character) and as
         <\diareverse{02B0}> for pre-aspiration (preceding the base
         character), so there is no consistent direction in which this
         diactritic should bind. Note that there is yet another homoglyph,
         namely the \textsc{combining latin small letter h} at \uni{036A}.
         We propose not to use this combining diacritical mark, but to
         standardize on Unicode modifier letters for all superscript
         letters in IPA.
